---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "README-"
)
```

# rong

_The rong approach to wrong inputs. Two (w)rongs make a right._

[![Build Status](https://travis-ci.org/egnha/rong.svg?branch=master)](https://travis-ci.org/egnha/rong)
[![codecov](https://codecov.io/gh/egnha/rong/branch/master/graph/badge.svg)](https://codecov.io/gh/egnha/rong)

> _rong_ is a complete reimplementation of
> [valaddin](https://github.com/egnha/valaddin) that supports tidyverse
> semantics.
>
> Since this is a major break in the API of the current
> [CRAN version](https://cran.r-project.org/package=valaddin) of valaddin (which
> appears to have no reverse dependencies), it make sense to start from a clean
> slate and distinguish this reconception with a new name.

## Overview

Dealing with invalid function inputs is a chronic pain for R users, given R’s 
weakly typed nature. _rong_ provides pain relief in the form of an adverb,
`firmly()`, that enables you to _transform_ an existing function into a function
with input validation checks, in a manner suitable for both programmatic and
interactive use.

Additionally, rong provides:
    
  * `fasten()`, to help you write cleaner and more explicit function
    declarations in your scripts, by providing a _functional operator_ that
    “fastens” a given set of input validations to functions (i.e., it 
    [curries](https://en.wikipedia.org/wiki/Currying) `firmly()`)
  
  * `validate()`, as syntactic sugar to validate _objects_, by applying input
    validation to the identity function
    
  * `loosely()`, to undo the application of input validation checks, at any
    time, by returning the original function
    
These functions support
[tidyverse semantics](https://rpubs.com/hadley/dplyr-programming) such as 
[quasiquotation](http://rlang.tidyverse.org/reference/quasiquotation.html) and 
[splicing](http://rlang.tidyverse.org/reference/quasiquotation.html), to provide
a flexible yet simple grammar for input validations.

## Installation

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("egnha/rong")
```

## Usage

To illustrate rong’s functional approach to input validation, consider the 
function that computes the barycentric coordinates of a point in the plane:

```{r}
bc <- function(x, y) {
  c(x, y, 1 - x - y)
}
```

### When validating inputs, think _function transformation_

Imagine applying `bc()` “firmly,” exactly as before, but with the assurance that
the inputs are indeed numeric. To enable this, transform `bc()` using
`firmly()`, relative to the validation specified by the predicate function
`is.numeric()`:

```{r error = TRUE, purl = FALSE}
library(rong)

bc2 <- firmly(bc, is.numeric)

bc2(.5, .2)

bc2(.5, ".2")
```

### Specify error messages that are context-aware

Using the string-interpolation syntax provided by the 
[glue](https://github.com/tidyverse/glue) package, make error messages more
informative, by taking into account the context of an error:

```{r error = TRUE, purl = FALSE}
bc3 <- firmly(bc, "{{.}} is not numeric (type: {typeof(.)})" := is.numeric)

bc3(.5i, ".2")
```

### Express input validations using tidyverse idioms

rong supports 
[quasiquotation](http://rlang.tidyverse.org/reference/quasiquotation.html) and 
[splicing](http://rlang.tidyverse.org/reference/quasiquotation.html) semantics 
for specifying input validation checks. Checks and (custom) error messages are 
captured as [quosures](http://rlang.tidyverse.org/reference/quosure.html), to 
ensure that validations, and their error reports, are hygienically evaluated in
the intended scope—transparently to the user.

```{r error = TRUE, purl = FALSE}
z <- 0
in_triangle <- vld_spec(
  "{{.}} is not positive (value is {.})" :=
    {isTRUE(. > !! z)}(x, y, 1 - x - y)
)

bc4 <- firmly(bc, is.numeric, !!! in_triangle)

bc4(.5, .2)

bc4(.5, .6)
```

This reads as follows:

  * `vld_spec()` encapsulates the condition that `x`, `y`, `1 - x - y` are
    positive, as a formula
    [definition](http://rlang.tidyverse.org/reference/quosures.html#details).
    The predicate itself is succinctly expressed using 
    [magrittr](https://github.com/tidyverse/magrittr)’s shorthand for anonymous 
    functions. The unquoting operator `!!` ensures that the _value_ of `z` is 
    “burned into” the check.
  
  * The additional condition that `(x, y)` lies in a triangle is imposed by
    splicing it in with the `!!!` operator.

### Use the same grammar to validate objects

Validating an object (say, a data frame) is nothing other than applying an
input-validated identity function to it. The function `validate()` provides a
shorthand for this.

```{r error = TRUE, purl = FALSE}
# All assumptions OK, mtcars returned invisibly
validate(mtcars,
         is.data.frame,
         chk_lt(100, nrow(.)),
         chk_has_names(c("mpg", "cyl")))

validate(mtcars,
         is.data.frame,
         chk_gt(100, nrow(.)),
         chk_has_name("cylinders"))
```

### Clarify code structure

Instead of writing

```{r}
bc_cluttered <- function(x, y) {
  if (!is.numeric(x) || length(x) != 1)
    stop("x is not a number")
  if (!is.numeric(y) || length(y) != 1)    
    stop("y is not a number")
  if (!isTRUE(x > 0))
    stop("x is not positive")
  if (!isTRUE(y > 0))
    stop("y is not in the upper-half plane")
  if (!isTRUE(1 - x - y > 0))
    stop("1 - x - y is not positive")

  c(x, y, 1 - x - y)
}
```

use `fasten()` to highlight the core logic, while keeping input assumptions in
sight:

```{r error = TRUE, purl = FALSE}
bc_clean <- fasten(
  "{{.}} is not a number" := {is.numeric(.) && length(.) == 1},
  "{{.}} is not positive" :=
    {isTRUE(. > 0)}(x, "y is not in the upper-half plane" := y, 1 - x - y)
)(
  function(x, y) {
    c(x, y, 1 - x - y)
  }
)

bc_clean(.5, .2)

bc_clean(c(.5, .5), -.2)
```

In addition to having cleaner code, you can:

  * reduce duplication, by using the splicing operator `!!!` to
    reuse common input validations
  
  * recover the underlying “lean” function, at any time, using `loosely()`:
    ```{r}
    loosely(bc_clean)
    ```

## Related packages

  * rong provides a basic set of predicate functions—prefixed `chk_` for 
    easy lookup—to specify common kinds of checks, e.g., type and property
    checks, comparisons, etc.
    
    To enrich rong’s vocabulary of predicate functions, use:
    
    * specialized collections of predicate functions, such as
      [assertive](https://bitbucket.org/richierocks/assertive),
      [assertthat](https://github.com/hadley/assertthat),
      [checkmate](https://github.com/mllg/checkmate)
    
    * [vetr](https://github.com/brodieG/vetr), which provides a concise
      declarative syntax to create custom predicate functions

  * Other non-functional approaches to input validation:
    [argufy](https://github.com/gaborcsardi/argufy),
    [ensurer](https://github.com/smbache/ensurer),
    [typeCheck](https://github.com/jimhester/typeCheck)
    
## Acknowledgement

rong makes essential use of the [rlang](https://github.com/tidyverse/rlang)
package by [Lionel Henry](https://github.com/lionel-) and
[Hadley Wickham](https://github.com/hadley), which provides the engine for
quasiquotation and expression capture. The
[glue](https://github.com/tidyverse/glue) package by
[Jim Hester](https://github.com/jimhester) enables string interpolation of error
messages.

## License

MIT Copyright © 2017 [Eugene Ha](https://github.com/egnha)
